Software for Automated Residual Plot Assessment: autovi and autovi.web

### 2024-11-18 autovi_paper.tex
compile the paper

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Software for Automated Residual Plot Assessment: autovi and autovi.web},
  pdfauthor={Weihao Li; Dianne Cook; Emi Tanaka; Susan VanderPlas; Klaus Ackermann},
  pdfkeywords={statistical software, statistical graphics, data
visualization, visual inference, computer vision, machine
learning, hypothesis testing, reression analysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Regression software is widely available today, but tools for effective
diagnostics are still lagging. Perhaps, one of the reasons is that it is
hard. Conventional tests, that would make the task easy, are often too
sensitive, which would result in adequate models being abandoned if the
analyst strictly adhered to the test decision. The recommended advice is
to have the analyst assess the strength of patterns in residual plots.
This requires human effort and also suffers from the potential for
inconsistent decisions from different analysts. Using a lineup protocol
can help to alleviate inconsistency, but requires even more human
effort. This is the type of task where a robot might be employed to do
the disagreeable work that currently requires a human. Here we describe
a new R package that includes a computer vision model for automated
assessment of residual plots, and an accompanying Shiny app for ease of
use. For a user-provided sample of residuals, it predicts a measure of
visual signal strength (VSS) and provides a suite of supporting
information to assist the analyst diagnose their model fit.
\end{abstract}

\keywords{statistical software; statistical graphics; data
visualization; visual inference; computer vision; machine
learning; hypothesis testing; reression analysis}

@@

Regression analysis is a fundamental statistical technique widely used
for modeling data from many fields. To diagnose the fit of a model it is
recommended that the residuals are plotted. If the fit is good, any
variation remaining should be noise, consistent with sampling from a
distribution specified by the error model. Deviations that might be
observed from a residual plot are non-normality, heteroscedasticity, and
other associations with the fitted values. In \citet{li2024plot}, we
demonstrated that visual methods for assessing residuals are
advantageous over conventional tests due to their reduced sensitivity to
minor departures. Building on this, \citet{li2024automated} introduced a
computer vision model aimed at alleviating the human effort required for
visual assessment of residual plots. The next step is to deliver the use
of the computer vision model to potential users, so that it can be
widely used to the benefit of the analytics community.

Software for regression analysis tools is widely available. The
Comprehensive R Archive Network (CRAN) \citep{hornik2012comprehensive}
hosts a vast array of packages, many of which provide tools to diagnose
models using residual plots. These packages can be broadly categorized
into three groups: general-purpose, enhanced visual diagnostics, and
visual diagnostics with statistical testing.

General-purpose regression analysis tools are the most widely used group
of packages. While not specifically designed for graphical diagnostics
of residuals in linear regression, they include this functionality as
part of a broader suite of statistical tools. For example, the
\texttt{stats} package \citep{stats} offers standard diagnostic plots
such as residuals vs.~fitted values, quantile-quantile (Q-Q) plots, and
residuals vs.~leverage plots. Additional packages like \texttt{jtools}
\citep{jtools}, \texttt{olsrr} \citep{olsrr}, \texttt{rockchalk}
\citep{rockchalk}, and \texttt{ggResidpanel} \citep{ggresidpanel}
provide similar graphical diagnostics, often with alternative aesthetics
or interactive features. All of these tools deliver the types of
diagnostic plots outlined in the classical text by
\citet{cook1982residuals}. However, as discussed in \citet{li2024plot},
relying solely on subjective assessments of these plots can lead to
issues, such as over-interpreting random patterns as model violations.

The second group consists of enhanced visual diagnostics tools, which
provide advanced aids for interpreting diagnostic plots. For instance,
\texttt{ecostats} \citep{warton_global_2023} incorporates simulation
envelopes into residual plots, while \texttt{DHARMa} \citep{dharma}
compares empirical quantiles (0.25, 0.5, and 0.75) of scaled residuals
to their theoretical counterparts. \texttt{DHARMa} is particularly
focused on detecting model violations such as heteroscedasticity,
incorrect functional forms, and issues specific to generalized linear
and mixed-effect models, like over/under-dispersion. It also includes
annotations from conventional tests, displayed as labels or text within
the plot, to help avoid misinterpretation.

The third group, visual diagnostics with statistical testing, offers
tools for formally testing patterns observed in diagnostic plots
\citep{buja2009statistical}. A common method for visual testing of
residual plots is the lineup protocol, where the true residual plot is
embedded within a set of null plots generated by simulating residuals
under the null hypothesis \(H_0\) that the regression model is correctly
specified. Observers are asked to identify the plot that appears most
different from the others. If a significant proportion of observers
correctly identify the true residual plot, it provides evidence against
\(H_0\), as the true residual plot should be indistinguishable from the
null plots if all residuals are generated by the same process, according
to \citet{buja2009statistical}. Packages in this group, such as
\texttt{nullabor} \citep{nullabor}, \texttt{HLMdiag}
\citep{loy2014hlmdiag}, and \texttt{regressinator}
\citep{regressinator}, enable users to compare observed residual plots
with samples from null distributions, helping to quantify the
significance of any detected patterns.

@@

The main purpose of \texttt{autovi} is to provide rejection decisions
and \(p\)-values for testing whether a regression model is correctly
specified. The package introduces a novel approach to automating
statistical analysis, particularly in the interpretation of residual
plots. The name \texttt{autovi} stands for automated visual inference.
While initially designed for linear regression residual diagnostics, it
has the potential to be extended to broader visual inference
applications, as we discuss in section
Section~\ref{sec-autovi-infrastructure}.

@@

An \texttt{autovi} checker can be initialized by supplying two primary
inputs, including a regression model object, such as an \texttt{lm}
object representing the result of a linear regression model, and a
trained computer vision model compatible with the \texttt{Keras}
\citep{chollet2015keras} Application Programming Interface (API), to the
\texttt{AUTO\_VI} class constructor \texttt{auto\_vi()}. The
\texttt{residual\_checker()} introduced in
Section~\ref{sec-autovi-usage} is a thin wrapper around
\texttt{auto\_vi()}, which will call \texttt{get\_keras\_model()} during
initialization. \texttt{get\_keras\_model()} is a function to download a
trained computer vision model (described in \citet{li2024automated})
from GitHub. ``vss\_phn\_32'' specifies a model that predicts VSS and is
trained on residuals with polynomial, heteroskedasticity, and
non-normality patterns (phn). More details about the hosted models will
be provided in Section~\ref{sec-trained-model-hosting}.

@@

\begin{verbatim}
[1] "/var/folders/61/bv7_1qzs20x6fjb2rsv7513r0000gn/T//RtmpfYtMdO/file37817b7b5b6.png"
\end{verbatim}

@@

\section{Availability}\label{availability}

The web interface is housed at \url{autoviweb.netlify.app}.

The the current version of \texttt{autovi} can be installed from CRAN,
and source code for both packages are available at
\url{https://github.com/TengMCing/}.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Software for Automated Residual Plot Assessment: autovi and autovi.web},
  pdfauthor={Weihao Li; Dianne Cook; Emi Tanaka; Susan VanderPlas; Klaus Ackermann},
  pdfkeywords={initial data analysis, statistical graphics, data
visualization, visual inference, computer vision, machine
learning, hypothesis testing, regression analysis},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Regression software is widely available, and there are many tools for
generating diagnostics and computing conventional residual tests.
However, the advice remains that the analyst should look at the residual
plot to check the fit. Perhaps, one of the reasons is that conventional
tests are too sensitive, strictly resulting in adequate models being
abandoned. Visually assess the strength of patterns in residual plots
requires human effort and suffers from the potential for inconsistent
decisions from different analysts. Using a lineup protocol, where the
residual plot is embedded among null plots, can help to alleviate
inconsistency, but requires even more human effort. This is the type of
task that in today's world might employ a robot to do the tedious work
for a human. Here we describe a new R package that includes a computer
vision model for automated assessment of residual plots, and an
accompanying Shiny app for ease of use. For a user-provided sample of
residuals, it predicts a measure of visual signal strength (VSS) and
provides a suite of supporting information to assist the analyst decide
on the appropriateness their model fit.
\end{abstract}

\keywords{initial data analysis; statistical graphics; data
visualization; visual inference; computer vision; machine
learning; hypothesis testing; regression analysis}

@@

Regression analysis is a widely used statistical modeling technique
widely for data in many fields. There are a vast array of software for
conducting regression modeling and generating diagnostics. The package
\texttt{lmtest} \citep{lmtest} provides a suite of conventional tests.
The \texttt{stats} package \citep{stats} offers standard diagnostic
plots such as residuals vs.~fitted values, quantile-quantile (Q-Q)
plots, and residuals vs.~leverage plots. Packages like \texttt{jtools}
\citep{jtools}, \texttt{olsrr} \citep{olsrr}, \texttt{rockchalk}
\citep{rockchalk}, and \texttt{ggResidpanel} \citep{ggresidpanel}
provide similar graphical diagnostics, often with alternative aesthetics
or interactive features. All of these tools deliver the types of
diagnostic plots outlined in the classical text by
\citet{cook1982residuals}. The \texttt{ecostats} package
\citep{warton_global_2023} incorporates simulation envelopes into
residual plots, while \texttt{DHARMa} \citep{dharma} compares empirical
quantiles (0.25, 0.5, and 0.75) of scaled residuals to their theoretical
counterparts. \texttt{DHARMa} is particularly focused on detecting model
violations such as heteroscedasticity, incorrect functional forms, and
issues specific to generalized linear and mixed-effect models, like
over/under-dispersion. It also includes conventional test annotations to
help avoid misinterpretation.

However relying solely on subjective assessments of these plots can lead
to issues, such as over-interpreting random patterns as model
violations. \citet{li2024plot} demonstrated that visual methods using
the lineup protocol \citep{buja2009statistical} for assessing residuals
are more useful, and also perform more practically than conventional
tests due to their reduced sensitivity to minor departures. Packages
such as \texttt{nullabor} \citep{nullabor}, \texttt{HLMdiag}
\citep{loy2014hlmdiag}, and \texttt{regressinator}
\citep{regressinator}, enable users to compare observed residual plots
with samples from null distributions, helping to quantify the
significance of any detected patterns. However, the

@@

The main purpose of \texttt{autovi} is to provide rejection decisions
and \(p\)-values for testing whether a regression model is correctly
specified. The package introduces a novel approach to automating
statistical analysis, particularly in the interpretation of residual
plots. The name \texttt{autovi} stands for automated visual inference.

There are two ways to access the package, directly using R or through a
web interface, \texttt{autovi.web}. The web interface has the advantage
that it can be used without installing Python, R and the relevant
packages locally.

\subsection{Why use it}\label{sec-why}

\textbf{?@fig-three-examples} shows three sets of plots of residuals
against fitted values. The simulated example in (a) might be interpreted
as a heteroscedastic pattern, however the automated reading would
predict this to have a visual signal strength (VSS) of 1.53, with a
corresponding \(p\)-value of 0.25. This means it is would be interpreted
as a good residual plot, that there is nothing in the data to indicate a
violation of model assumptions. Skewness in the predictor variables is
generating the apparent heteroscedasticity, where the smaller variance
in residuals at larger fitted values is due to smaller sample size only.
The Breusch-Pagan test \citep{breusch1979simple} for heteroscedasticity
would also not reject this as good residual plot.

The data in (b) is generated by fitting a linear model predicting
\texttt{mpg} based on \texttt{hp} using the \texttt{datasets::mtcars}.
It is a small data set, and there is a hint of nonlinear structure not
captured by the model. The automated plot reading would predict a VSS of
3.57, which has a \(p\)-value less than 0.05. That is, the nonlinear
structure is most likely real, and indicates a problem with the model.
The conventional test, a Ramsey Regression Equation Specification Error
Test (RESET) \citep{ramsey1969tests} would also strongly detect the
nonlinearity.

The third example is generated using the \texttt{surreal} package
\citep{surreal} where structured residuals are hidden in data, to be
revealed if the correct model is specified. Here a quote based on Tukey
is used as the residual structure ``visual summaries focus on unexpected
values''. The automated plot reading predicts the VSS to be 5.87, with a
\(p\)-value less than 0.05. This structure is blindingly obvious
visually, but a RESET test for nonlinear structure would not report a
problem. (It would be detected by a Breusch-Pagan for heteroscedasticity
and also Shapiro-Wilk test \citep{shapiro1965analysis} for
non-normality.)

{\centering \includegraphics[width=1\textwidth,height=\textheight]{autovi_paper_files/figure-pdf/three-examples-1.pdf}

\caption{Reading residual plots can be a difficult task, particular for
students new to statistical modeling. The autovi package makes it
easier. Here are three examples of residual plots, which may appear to
have structure. According to autovi, the visual signal strengths (VSS)
of these three examples are approximately (a) 1.53, (b) 3.57, (c) 5.87,
resulting in (b), (c) being significant violations of good residuals,
but (a) is consistent with a good residual plot.}

@@

An \texttt{autovi} checker can be initialized by supplying two primary
inputs, including a regression model object, such as an \texttt{lm}
object representing the result of a linear regression model, and a
trained computer vision model compatible with the \texttt{Keras}
\citep{chollet2015keras} Application Programming Interface (API), to the
\texttt{AUTO\_VI} class constructor \texttt{auto\_vi()}. The
\texttt{residual\_checker()} introduced in
Section~\ref{sec-autovi-usage} is a thin wrapper around
\texttt{auto\_vi()}, which will call \texttt{get\_keras\_model()} during
initialization. \texttt{get\_keras\_model()} is a function to download a
trained computer vision model (described in \citet{li2024automated})
from GitHub. ``vss\_phn\_32'' specifies a model that predicts VSS and is
trained on residuals with polynomial, heteroscedasticity, and
non-normality patterns (phn). More details about the hosted models will
be provided in Section~\ref{sec-trained-model-hosting}.

@@

\begin{verbatim}
[1] "/var/folders/61/bv7_1qzs20x6fjb2rsv7513r0000gn/T//Rtmp6Lce76/file135f86b4a7b69.png"
\end{verbatim}

@@

\section{Resources and Supplementary
Material}\label{resources-and-supplementary-material}

The the current version of \texttt{autovi} can be installed from CRAN,
and source code for both packages are available at
\url{https://github.com/TengMCing/autovi}. The web interface is
available from \url{autoviweb.netlify.app}.

These \texttt{R} packages were used for the work: \texttt{tidyverse}
\citep{tidyverse}, \texttt{lmtest} \citep{lmtest}, \texttt{kableExtra}
\citep{kableextra}, \texttt{patchwork} \citep{patchwork},
\texttt{rcartocolor} \citep{rcartocolor}, \texttt{glue} \citep{glue},
\texttt{here} \citep{here}, \texttt{magick} \citep{magick},
\texttt{yardstick} \citep{yardstick} and \texttt{reticulate}
\citep{reticulate}.
